Confirmation of Wide-Field Signatures in Redshifted 21~cm Power Spectra

### 2015-06-02 ms.tex
Make changes suggested by Judd, update references
From e734511:
============

In eor.bib,

(*) Add citation to Neben et al.(2015)
(*) Add citation to Wayth et al.(2015)

In ms.tex,

(*) Make changes suggested by Judd
(*) Make changes suggested by Danny
(*) Use updated contents of ms.bbl according to changes in eor.bib
above
(*) Emphasize increase in likelihood of signal detection in 3D k-space

We confirm our recent prediction of the ``pitchfork'' foreground signature in power spectra of high-redshift 21~cm measurements, wherein the interferometer is sensitive to large-scale structure on all baselines. This is due to the inherent response of a wide-field instrument and is characterized by enhanced power from foreground emission in Fourier modes adjacent to those considered to be most sensitive to the cosmological H~{\sc i} signal. In our recent paper, many signatures from the simulation which predicted this feature were validated against Murchison Widefield Array (MWA) data but this key element was at or below the noise level. In this paper, we improve the data sensitivity through coherent averaging of 12 independent snapshots with identical instrument settings, and provide the first confirmation of the prediction with a signal-noise ratio $>10$. This wide-field effect can by mitigated by careful antenna design which suppresses sensitivity towards the horizon. Simple models for antenna apertures proposed for future instruments such as the Hydrogen Epoch of Reionization Array and the Square Kilometre Array indicate they should suppress foreground contamination by $\sim 40$~dB relative to the MWA. However, detailed models are necessary to ensure the contamination stays below the expected cosmological signal.

@@

Our recent study \citep[][hereafter referred to as Paper~I]{thy15} used all-sky foreground and instrument models for the first time in order to simulate actual EoR experiments more accurately than previous studies. Surprisingly, we found that foreground emission outside the primary beam field of view caused the most significant contamination of Fourier modes considered most sensitive for detecting the cosmological H~{\sc i} signal in delay spectrum based analyses. This contamination is the result of the interplay between foreground emission, particularly diffuse Galactic emission, and the wide-field properties typical of EoR instruments. Our simulations predicted that delay spectra from the MWA and other experiments should exhibit a characteristic ``pitchfork'' appearance with local maxima near the horizon delay limits, in addition to at the primary lobe region.

@@

The defining characteristics of the {\it pitchfork} signature are understood as follows. The steep rise in subtended solid angle near the horizon for a fixed delay bin size significantly enhances the integrated emission near the horizon delay limits in wide-field measurements. This is found to be true for diffuse emission even on wide antenna spacings because their foreshortening towards the horizon makes them sensitive to large angular scales that match the inverse of their foreshortened lengths. We confirm this key feature using deeper MWA data.

@@

There is a remarkable agreement in broad morphology between the data and the model. However, some differences in the amplitude scales are noted. For instance, the emission near the positive horizon limit is higher in the data than in the model in both figures. We attribute such differences to uncertainties in the foreground model, the MWA tile power pattern, thermal fluctuations, and other uncertainties noted in Paper~I which limit a more thorough quantitative comparison between the model and the data. In fact, this lays further emphasis on the need for extending the footprint of surveys matching the frequency and angular resolution of observation such as the MWA Commissioning Survey \citep[MWACS;][]{hur14} and the Galactic and Extragalactic MWA Survey (GLEAM; Wayth et al. submitted) to cover the entire hemisphere, and a detailed characterization of the power pattern over the entire hemisphere.

@@

HERA will deploy 14~m dishes with 331 of them in a closely packed hexagonal array and another 21 as outliers. With the advantage of enormous redundancy and unprecedented sensitivity, it will address the following key questions -- what objects first lit up the Universe and reionized the neutral IGM? Over what redshift range did this occur? And how did the process proceed, leading to the large scale galaxy structure seen today? The fixed dishes will observe the sky drifting overhead. Based on simulations of uniformly illuminated circular disks, the response near the horizon is $\lesssim -40$~dB, $\gtrsim$~20~dB lower than that of the MWA. Tapering the reflectivity of the dish is being studied to further reduce this response. The responses of the actual dishes will be tested shortly.

With the SKA at low frequencies, deploying ``stations'' of $\sim$~35~m diameter that act as aperture arrays consisting of 256 pseudo-randomly placed vertical log-periodic dipole antennas is being considered\footnote{Documents at \url{https://www.skatelescope.org/key-documents/} on SKA re-baselining}$^,$\footnote{SKA memo ``Station Response and Imaging Performance of LFAA: 100 MHz to 600 MHz'' by Razavi-Ghods et al.}. The voltage beam of the station will be obtained by a phased addition of the dipole responses. Since the power pattern of a baseline consisting of two stations will be effectively a product of two voltage beams, it will tend to homogenize the fluctuations in the power patterns of individual stations. Even with unconstrained, and hence sub-optimal, realizations of these phased array stations, we estimate the typical horizon response of these beams to be $\gtrsim$~30--40~dB lower than at zenith, also $\sim 20$~dB lower than the MWA beam. Constrained optimization studies of antenna layouts of stations are under active study and could yield even better responses.

From these simple antenna aperture models, HERA and SKA should mitigate foreground contamination by $\gtrsim 40$~dB relative to the MWA. The precise beam responses and the resulting foreground contamination, especially near the horizon, will prove to be critical and timely inputs to the actual aperture designs of HERA dishes and SKA stations which are nearing their final stages.

@@

Using deeper MWA data, we have confirmed with high significance the earlier prediction that wide-field EoR measurements suffer significant foreground contamination from near the horizon. This has important implications for instrument design and data analysis of future instruments such as HERA and SKA. Precise modeling is thus required to gain a complete understanding of the characteristics of the cosmological signal and the foregrounds.

@@

For future work, we plan to extend our analysis to HERA. Based on Paper~I, such a dish will have a much desirable Fourier response from a foreground contamination viewpoint. One of our objectives is to forecast the per-baseline foreground contamination as a function of local sidereal time in order to tune the HERA observing strategy and data analysis to maximize sensitivity to the EoR signal.

@@

@@

>>>>

We confirm our recent prediction of the ``pitchfork'' foreground signature in power spectra of high-redshift 21~cm measurements, wherein the interferometer is sensitive to large-scale structure on all baselines. This is due to the inherent response of a wide-field instrument and is characterized by enhanced power from foreground emission in Fourier modes adjacent to those considered to be most sensitive to the cosmological H~{\sc i} signal. In our recent paper, many signatures from the simulation which predicted this feature were validated against Murchison Widefield Array (MWA) data but this key {\it pitchfork} signature was close to the noise level. In this paper, we improve the data sensitivity through coherent averaging of 12 independent snapshots with identical instrument settings, and provide the first confirmation of the prediction with a signal-noise ratio $>10$. This wide-field effect can by mitigated by careful antenna designs that suppress sensitivity near the horizon. Simple models for antenna apertures proposed for future instruments such as the Hydrogen Epoch of Reionization Array and the Square Kilometre Array indicate they should suppress foreground leakage from the {\it pitchfork} by $\sim 40$~dB relative to the MWA, and significantly increase the likelihood of cosmological signal detection in three-dimensional power spectrum.

@@

Our recent study \citep[][hereafter referred to as Paper~I]{thy15} used instrument and foreground models, for the first time with full sky coverage, in order to simulate actual EoR experiments more accurately than previous studies. Surprisingly, we found that foreground emission outside the primary beam field of view caused the most significant contamination of Fourier modes considered most sensitive for detecting the cosmological H~{\sc i} signal in delay spectrum based analyses. This contamination is the result of the interplay between foreground emission, particularly diffuse Galactic emission, and the wide-field properties typical of EoR instruments. Our simulations predicted that delay spectra from the MWA and other experiments should exhibit a characteristic ``pitchfork'' appearance with local maxima near the horizon delay limits, in addition to at the primary lobe region.

@@

The defining characteristics of the {\it pitchfork} signature are understood as follows. The steep rise in subtended solid angle near the horizon for a fixed delay bin size significantly enhances the integrated emission near the horizon delay limits in wide-field measurements. This is found to be true for diffuse emission even on wide antenna spacings because their foreshortening towards the horizon makes them sensitive to large angular scales that match the inverse of their foreshortened lengths. In the following sections, we present an observational confirmation of this effect.

@@

There is a remarkable agreement in broad morphology between the data and the model. However, some differences in the amplitude scales are noted. For instance, the emission near the positive horizon limit is higher in the data than in the model in both figures. We attribute such differences to uncertainties in the foreground model, the MWA tile power pattern, thermal fluctuations, and other uncertainties noted in Paper~I which limit a more thorough quantitative comparison between the model and the data. In fact, this lays further emphasis on the need for the following -- extending the footprint of surveys matching the frequency and angular resolution of observation such as the MWA Commissioning Survey \citep[MWACS;][]{hur14} and the Galactic and Extragalactic MWA Survey \citep[GLEAM;][]{way15} to cover the entire hemisphere; and, a detailed characterization of the power pattern over the entire hemisphere comprehensively extending studies such as those in \citet{neb15}.

@@

HERA will deploy 14~m dishes with 331 of them in a closely packed hexagonal array and another 21 as outliers. With the advantage of enormous redundancy and unprecedented sensitivity, it will address the following key questions -- what objects first lit up the Universe and reionized the neutral IGM? Over what redshift range did this occur? And how did the process proceed, leading to the large scale galaxy structure seen today? The fixed dishes will observe the sky drifting overhead. Based on simulations of uniformly illuminated circular disks, the response near the horizon is $\lesssim -40$~dB, which is over 20~dB lower than that of the MWA. Tapering the reflectivity of the dish is being studied to further reduce this response. The responses of the actual dishes will be tested shortly.

With the SKA at low frequencies, deploying ``stations'' of $\sim$~35~m diameter that act as aperture arrays consisting of 256 pseudo-randomly placed vertical log-periodic dipole antennas is being considered\footnote{Documents at \url{https://www.skatelescope.org/key-documents/} on SKA re-baselining}$^,$\footnote{SKA memo ``Station Response and Imaging Performance of LFAA: 100 MHz to 600 MHz'' by Razavi-Ghods et al.}. The voltage beam of the station will be obtained by a phased addition of the dipole responses. Since the power pattern of a baseline consisting of two stations will be effectively a product of two voltage beams, it will tend to homogenize the fluctuations in the power patterns of individual stations. Even with unconstrained, and hence sub-optimal, realizations of these phased array stations, we estimate the typical horizon response of these beams to be $-30$~dB to $-40$~dB compared to zenith, yielding a response that is at least 10~dB better than the MWA. Constrained optimization of antenna layouts of stations is under active study and could yield even better responses.

From these simple antenna aperture models, HERA and SKA should mitigate leakage from the {\it pitchfork} into the {\it EoR window} by $\sim 40$~dB relative to the MWA when expressed in units of delay power spectrum. This will significantly increase the likelihood of detecting the cosmological H~{\sc i} power spectrum in {\it voxels} in three-dimensional $k$-space. The precise beam responses and the resulting foreground leakage, especially from near the horizon, will prove to be critical and timely inputs to the actual aperture designs of HERA dishes and SKA stations which are nearing their final stages.

@@

Using deeper MWA data, we have confirmed with high significance the earlier prediction that wide-field EoR measurements suffer significant foreground contamination from near the horizon. This has important implications for instrument design and data analysis of future instruments such as HERA and SKA. Careful aperture designs that suppress the response near the horizon will significantly enhance the likelihood of detecting the cosmological signal in three-dimensional power spectrum. Precise modeling is thus required to gain a complete understanding of the characteristics of the cosmological signal and the foregrounds.

@@

For future work, we plan to extend our analysis to HERA. Based on Paper~I, a dish will have a much desirable Fourier response from a foreground contamination viewpoint. One of our objectives is to forecast the per-baseline foreground contamination as a function of local sidereal time in order to tune the HERA observing strategy and data analysis to maximize sensitivity to the EoR signal.

@@

\bibitem[{{Neben} {et~al.}(2015){Neben}, {Bradley}, {Hewitt}, {Bernardi},
  {Bowman}, {Briggs}, {Cappallo}, {Deshpande}, {Goeke}, {Greenhill},
  {Hazelton}, {Johnston-Hollitt}, {Kaplan}, {Lonsdale}, {McWhirter},
  {Mitchell}, {Morales}, {Morgan}, {Oberoi}, {Ord}, {Prabu}, {Udaya Shankar},
  {Srivani}, {Subrahmanyan}, {Tingay}, {Wayth}, {Webster}, {Williams}, \&
  {Williams}}]{neb15}
{Neben}, A.~R., {Bradley}, R.~F., {Hewitt}, J.~N., {et~al.} 2015, ArXiv
  e-prints, arXiv:1505.07114

@@

\bibitem[{{Wayth} {et~al.}(2015){Wayth}, {Lenc}, {Bell}, {Callingham},
  {Dwarakanath}, {Franzen}, {For}, {Gaensler}, {Hancock}, {Hindson},
  {Hurley-Walker}, {Jackson}, {Johnston-Hollitt}, {Kapinska}, {McKinley},
  {Morgan}, {Offringa}, {Procopio}, {Staveley-Smith}, {Wu}, {Zheng}, {Trott},
  {Bernardi}, {Bowman}, {Briggs}, {Cappallo}, {Corey}, {Deshpande}, {Emrich},
  {Goeke}, {Greenhill}, {Hazelton}, {Kaplan}, {Kasper}, {Kratzenberg},
  {Lonsdale}, {Lynch}, {McWhirter}, {Mitchell}, {Morales}, {Morgan}, {Oberoi},
  {Ord}, {Prabu}, {Rogers}, {Roshi}, {Udaya Shankar}, {Srivani},
  {Subrahmanyan}, {Tingay}, {Waterson}, {Webster}, {Whitney}, {Williams}, \&
  {Williams}}]{way15}
{Wayth}, R.~B., {Lenc}, E., {Bell}, M.~E., {et~al.} 2015, ArXiv e-prints,
  arXiv:1505.06041
